---
title: "Xenium Analysis Fig 4C_D SM41"
author: "Willa Kerkhoff"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}

library(Seurat)
library(future)
library(ggplot2)
library(arrow)
library(dplyr)
library(gridExtra)
library(ggplot2)

```

## Summary

This code allows easy recreation of Figure 4D and E in SM41 (purple), where we demonstrate 
the specificity and efficiency of RMacL3-01 for peripheral cortical layers in rhesus macaque. 

**Monkey**: SM41

**Injection, virus**: RMacL3-01, injected into dorsal bank of the principle sulcus

**Viral components**: HSP, GFP, 1st ranked L3 enhancer, dna barcode A5

**Date of Xenium run**: Feb. 21, 2025

**Section ID**: 294 LHem

**Xenium slide ID**: 0049095

**Xenium file name**: output-XETG00473__0049095__EUGENE__20250221__211144


## Analysis

We start by constructing a xenium object for SM41 using seurat.

```{r, message=FALSE, warnings=FALSE}

plan("multisession", workers = 10)
path <- "~/Desktop/20250221__211124__SM41_45_SM36_294_SM56_186/output-XETG00473__0049095__EUGENE__20250221__211144"

xenium.obj <- LoadXenium(path)

```

Outside of this code, ROIs across layers were outlined in SM41 using the lasso tool
in the Xenium Explorer Platform. The Explorer allows users to save the coordinates 
of their lassoed region as .csv files, which they can then use to subset their xenium
data for analysis. We have established ROIs for L1, L23, L4, and L56 in SM41 and saved
those in the xenium output file. The following code will create separate xenium objects
for each layer ROI.

```{r, message=FALSE, warnings=FALSE}

L1 <- read.csv("~/Desktop/20250221__211124__SM41_45_SM36_294_SM56_186/output-XETG00473__0049095__EUGENE__20250221__211144/EUGENElayer1_cells_stats.csv")
L23 <- read.csv("~/Desktop/20250221__211124__SM41_45_SM36_294_SM56_186/output-XETG00473__0049095__EUGENE__20250221__211144/EUGENElayer23_cells_stats.csv")
L4 <- read.csv("~/Desktop/20250221__211124__SM41_45_SM36_294_SM56_186/output-XETG00473__0049095__EUGENE__20250221__211144/EUGENElayer4_cells_stats.csv")
L56 <- read.csv("~/Desktop/20250221__211124__SM41_45_SM36_294_SM56_186/output-XETG00473__0049095__EUGENE__20250221__211144/EUGENElayer56_cells_stats.csv")

layer1 <- subset(xenium.obj, cells = unlist(L1))
layer23 <- subset(xenium.obj, cells = unlist(L23))
layer4 <- subset(xenium.obj, cells = unlist(L4))
layer56 <- subset(xenium.obj, cells = unlist(L56))

```

Now we use `FetchData()` to extract only the data we need for further analysis.
For each layer ROI, we want a table containing all cells that lists the transcript 
counts for the following features: `dna-barcode-A5-low`, `CUX2`, `RORB`, and 
`SLC17A7`. 

```{r}

L1 <- FetchData(object = layer1, vars = c("dna-barcode-A5-low","CUX2","RORB","SLC17A7"), layer = "counts")
L23 <- FetchData(object = layer23, vars = c("dna-barcode-A5-low","CUX2","RORB","SLC17A7"), layer = "counts")
L4 <- FetchData(object = layer4, vars = c("dna-barcode-A5-low","CUX2","RORB","SLC17A7"), layer = "counts")
L56 <- FetchData(object = layer56, vars = c("dna-barcode-A5-low","CUX2","RORB","SLC17A7"), layer = "counts")

```

From these tables, we will reproduce the findings reported in Figure 4D and 4E. We 
will begin by calculating specificity, shown in Figure 4D. 

**Specificity = # of A5+ cells in layer / all positive cells across all layers**

```{r}

# SM41 threshold for RMacL3_01+ classification
threshold <- 6

L1positive <- sum(L1$`dna-barcode-A5-low`>threshold) #count TRUE in logical vector
L23positive <- sum(L23$`dna-barcode-A5-low`>threshold)
L4positive <- sum(L4$`dna-barcode-A5-low`>threshold)
L56positive <- sum(L56$`dna-barcode-A5-low`>threshold)

A5positive <- L1positive + L23positive + L4positive + L56positive

L1specificity <- L1positive/A5positive
L23specificity <- L23positive/A5positive
L4specificity <- L4positive/A5positive
L56specificity <- L56positive/A5positive

Specificity <- data.frame(L1specificity,L23specificity,L4specificity,L56specificity)


```


To give us significance margins for specificity across layers, we're going to run
a permutation test to determine what level of specificity we would expect across
layers if RMacL3_01 infection was entirely random ($H_0$). 


```{r}

L1$layer <- rep(1, count(L1)$n)
L23$layer <- rep(2, count(L23)$n)
L4$layer <- rep(4, count(L4)$n)
L56$layer <- rep(5, count(L56)$n)

allLayers <- rbind(L1,L23,L4,L56)

set.seed(41)

n_perm <- 1000

perm_specificity <- rep(0,1)

for (i in 1:n_perm) {
  allLayers.perm <- transform(allLayers, layer = sample(layer))
  specificity.i <- allLayers.perm %>% group_by(layer) %>% 
  summarise(perm_specificity=sum(`dna-barcode-A5-low`>threshold)/A5positive,
            .groups = 'drop')
  perm_specificity <- rbind(perm_specificity,specificity.i)
}

perm_specificity <- subset(perm_specificity, layer > 0)

# what are the null means
permSpec.means <- perm_specificity %>% group_by(layer) %>% 
  summarise(mean= mean(perm_specificity),
            .groups = 'drop')

# define quantiles of interest
q = c(.5, .95)

# calculate quantiles by grouping variable
permSpec.quants <- perm_specificity %>%
                        group_by(layer) %>%
                        summarize(quant50 = quantile(perm_specificity, probs = q[1]), 
                                  quant95 = quantile(perm_specificity, probs = q[2]))
permSpec.quants

```


Grey-filled bars in 4D are the means calculated above, with error bars showing the
50% (quant50) and 95% (quant95) bounds for the null distribution. Importantly, the 
bounds generated by this code will vary *slightly* from what is reported in Figure 4D, 
as this analysis relies on a shuffle based test that will always include some variation. 

At this point, we are going to switch gears and focus on efficiency rather than 
specificity. The following analysis is relevant for Figure 4E SM41 (purple).

**Efficiency = # of A5+ cells in layer / all cells in layer**

```{r}

L1efficiency <- L1positive/count(L1)$n
L23efficiency <- L23positive/count(L23)$n
L4efficiency <- L4positive/count(L4)$n
L56efficiency <- L56positive/count(L56)$n

Efficiency <- data.frame(L1efficiency,L23efficiency,L4efficiency,L56efficiency)

```


Now for significance. We'll be calculating the standard error of a proportion for 
our error bars and running Chi-squared tests to assess pair-wise significance between
efficiency in L23 and each other layer.

```{r}

# se = sqrt(p*(1-p)/n)

L1er <- sqrt(L1efficiency*(1-L1efficiency)/count(L1)$n)
L23er <- sqrt(L23efficiency*(1-L23efficiency)/count(L23)$n)
L4er <- sqrt(L4efficiency*(1-L4efficiency)/count(L4)$n)
L56er <- sqrt(L56efficiency*(1-L56efficiency)/count(L56)$n)

Eff.er <- data.frame(L1er,L23er,L4er,L56er)

# proportion test between L23 and each of the other layers (L1, L4, L56)

L23.L1test <- prop.test(x = c(L1positive, L23positive), n = c(count(L1)$n, count(L23)$n))
L23.L1test

L23.L4test <- prop.test(x = c(L4positive, L23positive), n = c(count(L4)$n, count(L23)$n))
L23.L4test

L23.L56test <- prop.test(x = c(L56positive, L23positive), n = c(count(L56)$n, count(L23)$n))
L23.L56test

```




